---
pagetitle: "Spatial Data Science"
author: "Dr Thiyanga S. Talagala"
date: "18 February 2024"
host: "MSc in Data Science and Intelligence"
institute: "University of Sri Jayewardenepura"
strip-comments: true
format: 
  revealjs:
    slide-number: true
    multiplex: true
    theme: assets/monash.scss
    show-slide-number: all
    controls: true
    width: 1280
    height: 720
    incremental: true
    auto-stretch: false
    css: [assets/custom.css, assets/pacman.css, assets/tachyons-addon.css]
    include-after: |
      <link rel="stylesheet" href="assets/syntax-highlight.css">
      <link rel="stylesheet" href="assets/fontawesome-free-6.1.1-web/css/all.min.css">
      <script src="assets/fontawesome-free-6.1.1-web/js/all.min.js"></script>
callout-icon: false
output-file: index.html
execute:
  echo: true
---

## {#title-slide background="images/book-bg.jpg"}

```{r setup, include = FALSE}
library(tidyverse)

rotating_text <- function(x, align = "center") {
  glue::glue('
<div style="overflow: hidden; height: 1.2em;">
<ul class="content__container__list {align}" style="text-align: {align}">
<li class="content__item">{x[1]}</li>
<li class="content__item">{x[2]}</li>
<li class="content__item">{x[3]}</li>
<li class="content__item">{x[4]}</li>
</ul>
</div>')
}

fa_list <- function(x, incremental = FALSE) {
  icons <- names(x)
  fragment <- ifelse(incremental, "fragment", "")
  items <- glue::glue('<li class="{fragment}"><span class="fa-li"><i class="{icons}"></i></span> {x}</li>')
  paste('<ul class="fa-ul">', 
        paste(items, collapse = "\n"),
        "</ul>", sep = "\n")
}

```

::: title-box

<h2>`r rmarkdown::metadata$pagetitle`</h2>

👩🏻‍💻 [`r rmarkdown::metadata$author` @ `r rmarkdown::metadata$institute`]{.author}  


`r rotating_text(c('<i class="fa-solid fa-envelope"></i> ttalagala@sjp.ac.lk', '<i class="fa-brands fa-twitter"></i> @thiyangt', '<i class="fa-brands fa-github"></i> github.com/thiyangt', '<i class="fa-solid fa-globe"></i> https://thiyanga.netlify.app/'))`


:::



## Outline  {background-color=""}

- Introduction

- Three types of spatial processes

- Isotropy

- Spatial kriging

## Introduction

A spatial stochastic process is a family of random variables $$\{Z(s): s \in D\}.$$

Here, 

$D$ is the spatial domain in which observations could be made and 

$Z(s)$ is a random variable representing the quantity  that you measured at location $s$.

## Daily temperature in Canada

Locations of Canadian weather stations: $s_i$

```{r}
library(sf)
library(geoFourierFDA)
library(rnaturalearth)
canada$m_coord
```


Example: $s_1 = (50.25, -104.39)$

## Daily temperature in Canada

Observed daily temperature value: $z(s_i)$

```{r}
canada$m_data[1, ]
```

Example: 

$$s_1 = (50.25, -104.39)$$
$$z(s_1) = -15.7$$

## Visualization

```{r}
df <- data.frame(canada$m_coord)
df
```

## Visualization (cont.)

```{r}
df$location <- attr(canada$m_coord, "dimnames")[[1]]
df
```

## Visualisation (cont.)

```{r}
df <- st_as_sf(df, coords=c("W.longitude", "N.latitude"))
df
```

## Visualisation (cont.)

```{r}
st_crs(df) <- 4326
df
```

## Add temperature data

```{r}
df$temp <- canada$m_data[1, ]
df
```

## Getting map data

```{r}
map <- rnaturalearth::ne_states("Canada", returnclass = "sf")
map
```

## Visualisation of the spatial domain $D$

This represents $D$ - spatial domain (geographical region which the observations could be made)

```{r}
ggplot(map) + geom_sf()
```

::: {.f3 .monash-gray80 style="line-height:0.9!important;"}

A spatial stochastic process is a family of random variables $$\{Z(s): s \in D\}.$$

:::


## Visualisation of the spatial domain ($D$), and $z(s_i)$ values


```{r}
library(viridis)
ggplot(map) + geom_sf() + geom_sf(data=df, aes(color = temp), size=3) + scale_color_viridis()
```



## Three types of spatial processes

1. Geostatistical processes: 

   In theory the process could be observed at **infinitely many locations** in $S \in D$.
   
   Example: Air temperature in Sri Lanka

2. Areal process
    
      Here $D$ is divided into on-overlapping sub regions such as $D = {B_1, B_2,...B_n}$ where $B_i$ is the $i^{th}$ sub region.
  
    
    Example: Total number of dengue cases in each district in Sri Lanka
 
3. Point process

    The process can be observed only at specific locations in $S \in D$.
  
    Example: Tree volume of sandalwood trees in Sri Lanka



## Tobler’s First Law of Geography

Everything is related to everything else, but near things are more related than distant things.

```{r}
library(viridis)
ggplot(map) + geom_sf() +
  geom_sf(data=df, aes(color = temp), size=3) +
  scale_color_viridis()
```


## Weakly Stationary


## Strictly Stationary


##

:::: {.columns}

::: {.column width="50%"}

### Isotropy

direction-independent

```{r}
#| echo: false
#| fig-width: 15
#| fig-height: 15
set.seed(99999)
large_number <- 1e5
x <- rnorm(large_number)
y <- rnorm(large_number)
z <- cbind(x, y)
z <- data.frame(z)

ggplot(z) +
  geom_density2d_filled(aes(x = x, y = y)) +
   theme(legend.position = "none") +
  coord_fixed()

```

:::

::: {.column width="50%"}

### Anisotropic

direction-independent

```{r}
#| echo: false
#| fig-width: 15
#| fig-height: 15
A <- matrix(c(1, 0,
              0.9, sqrt(1 - 0.8^2)), byrow = TRUE, nrow = 2, ncol = 2)
x <- t(A %*% t(z)) 
colnames(x) <- c('x', 'y') # forces the aspect ratio to be 1:1 so you see the ellipses!
ggplot(data.frame(x)) +
  geom_density2d_filled(aes(x, y)) +
  theme(legend.position = "none") +
  coord_fixed()

```

:::

::::


##

1. What is isotropic?
  
   The correlation between any two observations depends only on the **distance between those locations** and not on their relative direction (directionally invariant).


2. Why is isotropic important in spatial data analysis?

   This assumption simplifies analysis and modeling processes.



## Spatial Interpolation

Spatial interpolation is the activity of estimating values of spatially continuous variables (fields) for spatial locations where they have not been observed, based on observations.

### Spatial Interpolation Methods

- Inverse Distance Weighting (IDW): It estimates values for unmeasured locations by considering the values and distances of nearby measured points.

- Kriging: A geostatistical technique that provides optimal predictions and quantifies uncertainty by analyzing spatial autocorrelation.


- Spline Interpolation: It fits a smooth curve through measured points to estimate values at unmeasured locations.

- Nearest-neighbor interpolation: It assigns the value of the closest measured point to the unmeasured location.

## Statial Interpolation Methods (cont.)

- Natural Neighbor: Similar to nearest neighbor but considers multiple nearby points and their influence.

- Radial Basis Function (RBF): It constructs a continuous function that satisfies given data values and radial symmetry properties.

- Deep Learning Model for Spatial Interpolation 

- Image-driven spatial interpolation

- Bayesian deep learning approaches

- Ensemble approaches

## Inverse Distance Weighting (IDW)

- Nonparametric interpolation method

- Based on weighted averages of the observations. 

- Weights are calculated based on the inverse of the distances between the known data points and the location where the estimation is being made

- Relies on Tobler’s first law of geography

## Inverse Distance Weighting (IDW)

Suppose $s_0$ is the point that we need to interpolate. Then,

$$\hat{Z}(s_0) = \frac{\sum_{i=1}^{n}w(s_i)Z(s_i)}{\sum_{i=1}^nw(s_i)}$$ and $w_i$ is defined as 

$$w_i = \frac{1}{||s_i - s_o||^p}$$

- We can tune the value of $p$ using a cross validation method

- We can limit the amount of neighbours to take into consideration to speed up the computations.

## Your turn

:::: {.columns}

::: {.column width="40%"}

```{r, comment=NA, echo=FALSE}
location <- 1:5
x <- c(2, 3, 9, 6, 5)
y <- c(2, 7, 9, 5, 3)
z <- c(10, 11, 15, 9, 8)
#d_5_5 <- sqrt((x - 5)^2 + (y-5)^2)
df1 <- data.frame(x=x, y=y, z=z)
knitr::kable(df1)
```

:::

::: {.column width="60%"}

```{r, comment=NA, echo=FALSE}
ggplot(df1, aes(x=x, y=y, col=z)) + geom_point(size=3) + coord_fixed()
```

:::

::::

We want to find z value at x=5, y=5.

## Answer

```{r, comment=NA, echo=FALSE}
location <- 1:5
x <- c(2, 3, 9, 6, 5)
y <- c(2, 7, 9, 5, 3)
z <- c(10, 11, 15, 9, 8)
d_5_5 <- sqrt((x - 5)^2 + (y-5)^2)
df1 <- data.frame(x=x, y=y, z=z, dis_to_55=d_5_5)
df1$inv_dis <- 1/d_5_5
df1$weight <- df1$inv_dis/sum(df1$inv_dis)
knitr::kable(df1)
```

Predicted value at (5, 5)

$z(5, 5) = 10\times 0.104 + 11 \times 0.156 + 15 \times 0.078 + \\ 9 \times 0.441 + 8 \times 0.220$

## IDW - Application


<!--https://www.paulamoraga.com/book-spatial/spatial-interpolation-methods.html-->

<!--https://www.geo.fu-berlin.de/en/v/soga-r/Advances-statistics/Geostatistics/Inverse-Distance-Weighting-IDW/IDW-Interpolation-of-Weather-Data/index.html-->

<!--https://bookdown.org/igisc/EnvDataSci/spatial-interpolation.html-->

Question: Interpolate Mean $NO_2$
 concentrations in air  in Germany, in 2017

Step 1: Load necessary packages

```{r, warning=FALSE, message=FALSE}
library(gstat)
library(tidyverse)
```

Step 2: Read data

```{r}
no2 <- read_csv(system.file("external/no2.csv", 
    package = "gstat"), show_col_types = FALSE)
no2
```



##

```{r}
head(data.frame(no2), 3)
```

## Create a Coordinate Reference System (CRS) Object

- st_crs: used to create a coordinate reference system (CRS) object.

- `EPSG:32632` is a specific CRS identifier. In this case, it refers to a CRS known as "WGS 84 / UTM zone 32N". This CRS is commonly used for mapping and spatial analysis in Europe, particularly in the northern regions.

- The following line of code creates a CRS object named crs representing the coordinate reference system "WGS 84 / UTM zone 32N".



```{r}
crs <- st_crs("EPSG:32632")
crs
```



## convert no2 to spatial object

```{r}
no2.sf <- st_as_sf(no2, crs = "OGC:CRS84", coords = 
    c("station_longitude_deg", "station_latitude_deg")) |>
    st_transform(crs) 
no2.sf
```

## EPSG: 32632 vs OGC: CRS84

- EPSG:32632
  
  Commonly used for regional mapping and spatial analysis in Europe, particularly in the northern regions. 

- OGC:CRS84
  
  Often used in web mapping applications and geospatial data exchange due to its simplicity and widespread support. 
  

-  EPSG:32632 is a regional projected CRS optimized for mapping specific areas with minimal distortion, while OGC:CRS84 is a global geographic CRS commonly used for general-purpose mapping and data exchange. 

## Load country boundary

```{r}

map <- read_sf("de_nuts1.gpkg") |> st_transform(crs) 
map
```

::: {.f3 .monash-gray80 style="line-height:0.9!important;"}

Data source: https://github.com/edzer/sdsr/tree/main/data 
 (select de_nuts1.gpkg)

:::

## Plot map

```{r}
ggplot() + geom_sf(data = map)
```

## Add observed values to the map

```{r}
ggplot() + geom_sf(data = map) + 
geom_sf(data = no2.sf, mapping = aes(col = NO2)) + 
  scale_color_viridis()
```

## IDW 
Step 1: Create a regular grid ($10km \times 10km$)

```{r, warning=FALSE}
library(stars)
grid <- st_bbox(map) |>
  st_as_stars(dx = 10000) |> st_crop(map) 
grid
```

- `st_bbox(map)`: Calculate the bounding box of the spatial object `map`.

##

```{r, warning=FALSE}
#| eval: false
library(stars)
grid <- st_bbox(map) |>
  st_as_stars(dx = 10000) |> st_crop(map) 
grid
```

- `st_as_stars(dx = 10000)`: Convert the bounding box obtained from `st_bbox(map)` into a raster object using the stars package. The `dx` parameter sets the cell size (in the x-direction) for the resulting raster, with a value of 10000 specified here.


- `st_crop(map):` This function crops the raster grid obtained from st_as_stars(dx = 10000) to match the extent of the map object. This ensures that the resulting grid covers only the area within the spatial extent of the map object.

## IDW 

Step 2: Interpolation

```{r}
library(gstat)
interpolated.values <- idw(NO2~1, no2.sf, grid)
interpolated.values 
```

## IDW

Step 3: Plot values

```{r}
ggplot() + geom_stars(data = interpolated.values, 
                      aes(fill = var1.pred, x = x, y = y)) + 
    geom_sf(data = st_cast(map, "MULTILINESTRING")) + 
    geom_sf(data = no2.sf, col="red") + scale_fill_viridis() 
```

## Comparison

```{r}
library(patchwork)
p1 <- ggplot() + geom_sf(data = map) +  geom_sf(data = no2.sf, mapping = aes(col = NO2)) + scale_color_viridis()
p2 <- ggplot() + geom_stars(data = interpolated.values, 
                      aes(fill = var1.pred, x = x, y = y)) + geom_sf(data = st_cast(map, "MULTILINESTRING")) + geom_sf(data = no2.sf, col="red") + scale_fill_viridis() 
p1|p2
```



## Variogram Modelling and Krigging

Geostatistical processes are generally modelled using additive decomposition:

$$Z(s) = \mu(s) + \epsilon(s)$$

Where,

$\mu(s)$ - large-scale variation/ spatial-trend/ spatially varying mean that captures the long range (nonstationary) spatial variation

$\epsilon(s)$ - small-scale variation/ innovations/ spatial geostatistical error

- To fully capture the characteristics of the geostatistical process, we need to model both $\mu(s)$ and $\epsilon(s)$.

## Overall modeeling approach

$$Z(s) = \mu(s) + \epsilon(s)$$

First we model $\mu(s)$ using OLS or some other trend model  via covariates and the residuals are used to model small-scale variation. 


## Modelling Large Scale Variation

- Regression

- Machine-learning models


## Modelling Small Scale Variation

Let's learn some theories we need to model small scale variations.

## Investigate Spatial Autocorrelation Using Empirical Variogram/ Variogram cloud

$$\gamma(s, t) = \frac{1}{2}Var\{Z(s_i) - Z(s_j)\}$$
An estimator for this is

$$\gamma(s, t) = \frac{1}{2}\{Z(s_i) - Z(s_j)\}^2$$

Proof: In-class

## Variogram Cloud/ Empirical Variogram

```{r}
vcloud <- variogram(NO2~1, no2.sf, cloud=TRUE)
vcloud
```

## Illustration: `gamma` value calculation

```{r}
right <- no2[1,]$NO2
right
left <- no2[3,]$NO2
left
(right-left)^2/2
```

## Illustration: `dist` value calculation

Your turn


## Variogram Cloud: Method 1

```{r}
ggplot(data=vcloud, aes(x=dist, y=gamma)) + geom_point()
```


## Variogram Cloud: Method 2

```{r}
plot(vcloud)
```

## Binned Empirical Semi-Variogram/ Sample variogram/ Sample Experimental Variogram

$$\hat{\gamma}(h) = \frac{1}{N_h}\sum_{i=1}^{N_h}(Z(s_i)-Z(s_i+h))^2$$
where 

$h$ is the spatial distance between observations

$N_h$ is the number of pairs that are within $h$ distance.


##  Binned Empirical Semi-Variogram/ Sample variogram/ Sample Experimental Variogram

```{r}
v1 <- variogram(NO2~1, no2.sf)
v1
```

`cutoff`: one-third of the length of the bounding box diagonal

`width`: cutoff divided by 15

## Plot Sample Experimental Variogram

```{r}
plot(v1, plot.numbers = TRUE, xlab = "distance h [m]",
     ylab = expression(gamma(h)),
     xlim = c(0, 1.055 * max(v1$dist)))
```

## Change default values

```{r}
v2 <- variogram(NO2~1, no2.sf, cutoff = 100000, width = 10000)
plot(v2, plot.numbers = TRUE, xlab = "distance h [m]",
     ylab = expression(gamma(h)),
     xlim = c(0, 1.055 * max(v2$dist)))
```

## Fitting variogram models

```{r}
vgm()
```

## Variogram models available in R’s gstat package

```{r}
show.vgms(par.strip.text=list(cex=0.75))
```

## Variogram

Resources: https://surferhelp.goldensoftware.com/vario/variogram_model.htm

https://csegrecorder.com/articles/view/the-variogram-basics-a-visual-intro-to-useful-geostatistical-concepts

## Variogram modelling

```{r}
v.m <- fit.variogram(v1, vgm(psill=20, model = "Exp", range = 20000, nugget = 1))
```

We need to provide partial sill, range, and nugget


## Visualize the fit

```{r}
plot(v1, v.m)
```

## Assess isotropy assumption

```{r}
v1.ani <- variogram(NO2~1, alpha = c(0, 45, 90, 135), no2.sf)
v.m <- fit.variogram(v1, vgm(psill=20, model = "Exp", range = 20000, nugget = 1))
plot(v1.ani, v.m)
```

##

```{r}
plot(v1.ani)
```

## Addressing anisotropy parameters

```{r}
v1.ani <- variogram(NO2~1, alpha = c(0, 45, 90, 135), no2.sf)
fit.ani <- vgm(psill=20, model = "Exp", range = 25000, nugget = 3, anis = c(30, 10, 0, 0.5, 0.3))
plot(v1.ani, fit.ani)
```

Your turn: Fit a suitable model (I will proceed with `v.m` )

## Krigging

1. Simple kriging

2. Ordinary kriging

3. Universal kriging

4. Regression kriging

5. Others: Staratified Kriging, Indicator Kriging, Block kriging

## Ordinary Kriging

```{r}
v1 <- variogram(NO2~1, no2.sf)
v.m <- fit.variogram(v1, vgm(psill=20, model = "Exp", range = 20000, nugget = 1))
krigOK <- krige(NO2~1, no2.sf, grid, v.m)
krigOK
```

## Visualise kriging predictions

```{r}
ggplot() + geom_stars(data = krigOK, aes(fill = var1.pred, x = x, y = y)) + 
    xlab(NULL) + ylab(NULL) +
    geom_sf(data = st_cast(map, "MULTILINESTRING")) + 
    geom_sf(data = no2.sf) +
    coord_sf(lims_method = "geometry_bbox") + scale_fill_viridis()
```

## Validation

```{r}
krig.ok.cv <- krige.cv(NO2~1, no2.sf, v.m)
krig.ok.cv 
```

### RMSE

```{r}
round(sqrt(mean(krig.ok.cv$residual^2)), 2)
```

## Regression Kriging

There is a trend in $\mu(s)$. This trend is estimated using covariates other than coordinates.

```{r}
plot( no2.sf$station_altitude, no2.sf$NO2,)
```

## Sample Variogram

```{r}
v.rk <- variogram(NO2~station_altitude, no2.sf)
plot(v.rk)
```

## Fit Variogram

```{r}
v.m.rk <- fit.variogram(v.rk, vgm(psill=15, model = "Exp", range = 20000, nugget = 1))
plot(v1, v.m)
```

## Assess isotropy assumption

```{r}
v1.ani.rk <- variogram(NO2~station_altitude, alpha = c(0, 45, 90, 135), no2.sf)
v.m.rk <- fit.variogram(v1.ani.rk, vgm(psill=15, model = "Exp", range = 20000, nugget = 1))
plot(v1.ani.rk, v.m.rk)
```

## Regression Kriging

Your turn: Complete regression kriging

## Using covariate

```r
vr <- variogram(NO2~sqrt(pop_dens), no2.sf)
vr.m <- fit.variogram(vr, vgm(1, "Exp", 50000, 1))
```

## Your turn

<!--https://www.paulamoraga.com/book-spatial/kriging.html-->

```{r}
library(sp)
library(gstat)
library(sf)
library(mapview)

data(meuse)
data(meuse.grid)

meuse <- st_as_sf(meuse, coords = c("x", "y"), crs = 28992)
mapview(meuse, zcol = "zinc",  map.types = "CartoDB.Voyager")

meuse.grid <- st_as_sf(meuse.grid, coords = c("x", "y"),
                       crs = 28992)
mapview(meuse.grid,  map.types = "CartoDB.Voyager")
```


## Perform EDA

```{r}
meuse
```


## Variogram Modelling with Pyhon

[Click here to view the tutorial](https://scikit-gstat.readthedocs.io/en/latest/auto_examples/tutorial_01_getting_started.html)

## Variogram Interpretations

[Resources 1](https://csegrecorder.com/articles/view/the-variogram-basics-a-visual-intro-to-useful-geostatistical-concepts)

[Resources 2](https://www.semanticscholar.org/paper/Methodology-for-Variogram-Interpretation-and-for-Gringarten-Deutsch/c12c97b6da9a7c1ee163c86cb5e669eed3dc3976)
